Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-04 source/paper.tex
some transport terms

The finite element discretisation is formed by choosing appropriate
representations, or function spaces, for the prognostic variables,
multiplying by test functions from these function spaces and
integrating over the domain. In addition to this, we integrate by
parts when the interelement continuity of the basis functions is not
sufficient to define the result of applying the differential operator
directly.

In line with the theory outlined in the introduction, we always choose
the velocity $\MM{u} \in \mathbb{V}^{HDiv}$ with $\mathbb{V}^{HDiv}
\subset H(\text{div})$. **mention bcs here! The field whose gradient
appears in the `pressure' gradient term is chosen to be in the space
$\mathbb{V}^{L^2}$ with $\mathbb{V}^{L^2} \subset L_2$, or, in the
case of the more complex nonlinear pressure gradient term for the
compressible Euler equations, it is a function of a variable in this
space. The compressible Boussinesq and compressible Euler equations
have a third prognostic variable, either buoyancy or potential
temperature, which is chosen to be in the space $\mathbb{V}_\theta$
which has degrees of freedom that are collocated with those of the
vertical velocity, mimicking the Charney-Phillips staggering as
discussed in **. These choices are governed by the properties required
of discretisation of the linearised equation sets (as discussed in
section ** and other refs). To solve the nonlinear equation sets we
need to introduce stable and accurate discretisations of the transport
terms and the nonlinear pressure gradient term in the compressible
Euler equation set. In addition, some test cases require viscous and
diffusive terms to generate converged solutions \ref{}. In the
following subsections we illustrate a non-exhaustive selection of the
options available in Gusto.

@@

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x, \quad \forall \MM{w} \in \mathbb{V}_u,
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element facets and on exterior boundaries we set no normal flow
boundary conditions**. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integration by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in \mathbb{V}_\rho$ and $\theta \in \mathbb{V}_\theta$ so is
discontinuous between elements, but then integrating by parts
transfers the differential operator to $\MM{w}\theta$ which is
discontinuous in the horizontal direction. Following \ref{} we integrate by parts
in each element $e$, giving

\begin{equation}
  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\hat{\theta}\MM{w}\cdot\MM{n}\hat{\Pi}\diff S,
\end{equation}
where $\partial e$ denotes the element facets, $\MM{n}$ is the outward
pointing normal to the facet and $\hat{q}$ denotes the value of a
variable $q$ on the facet, which for discontinuous variables does not
have a unique definition. Summing over the domain gives

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
as in \citet{}, to take the average value of $\Pi$ across the facets,
$\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each
side of the facet. The brackets $\jump{q} = q^+-q^-$ denote the jump
in $q$ across the facet and the integral over the horizontal facets
vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous
across these facets so the jump is zero.

\subsection{Transport terms}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable. The spatial order of the
finite element spaces also imposes requirements on the choice of
transport scheme. In particular, when the lowest order space
$\mathbb{V}_p$ is chosen to be DG0, we need to use the recovered
transport scheme described in \citet{bendall2019recovered} to achieve
second order convergence overall. Here we describe the default
transport schemes used in Gusto for higher order function spaces and
refer the reader to \citet{bendall2019recovered} for details of the
recovered schemes, which are also implemented in Gusto.

>>>>

The finite element discretisation is formed by choosing appropriate
representations, or function spaces, for the prognostic variables,
multiplying by test functions from these function spaces and
integrating over the domain. In addition to this, we integrate by
parts when the interelement continuity of the basis functions is not
sufficient to define the result of applying the differential operator
directly. For the domains considered in this paper, the only boundary
conditions we use are those of no normal flow, $\MM{u}\cdot\MM{n}=0$
where $\MM{n}$ is the outward pointing normal to the boundary. This
means that the test function for the velocity must also be chosen from
a function space that has this boundary condition.

In line with the theory outlined in the introduction, we always choose
the velocity $\MM{u} \in \ring{\mathbb{V}}^{HDiv}$ with
$\ring{\mathbb{V}}^{HDiv} \subset \mathbb{V}^{HDiv} \subset
H(\text{div})$ and, as mentioned above, the $\ring{}$ indicates that
the functions in $\ring{\mathbb{V}}^{HDiv}$ satisfy the no normal flow
boundary conditions. The field whose gradient appears in the
`pressure' gradient term is chosen to be in the space
$\mathbb{V}^{L^2}$ with $\mathbb{V}^{L^2} \subset L_2$, or, in the
case of the more complex nonlinear pressure gradient term for the
compressible Euler equations, it is a function of a variable in this
space. The compressible Boussinesq and compressible Euler equations
have a third prognostic variable, either buoyancy or potential
temperature, which is chosen to be in the space $\mathbb{V}_\theta$
which has degrees of freedom that are collocated with those of the
vertical velocity, mimicking the Charney-Phillips staggering as
discussed in **. These choices are governed by the properties required
of discretisation of the linearised equation sets (as discussed in
section ** and other refs). To solve the nonlinear equation sets we
need to introduce stable and accurate discretisations of the transport
terms and the nonlinear pressure gradient term in the compressible
Euler equation set. In addition, some test cases require viscous and
diffusive terms to generate converged solutions \ref{}. In the
following subsections we illustrate a non-exhaustive selection of the
options available in Gusto.

@@

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element facets and on exterior boundaries we set no normal flow
boundary conditions. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integration by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in \mathbb{V}^{L^2}$ and $\theta \in \mathbb{V}_\theta$ so is
discontinuous between elements, but then integrating by parts
transfers the differential operator to $\MM{w}\theta$ which is
discontinuous in the horizontal direction. Following \ref{} we integrate by parts
in each element $e$, giving

\begin{equation}
  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\hat{\theta}\MM{w}\cdot\MM{n}\hat{\Pi}\diff S, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
\end{equation}
where $\partial e$ denotes the element facets, $\MM{n}$ is the outward
pointing normal to the facet and $\hat{q}$ denotes the value of a
variable $q$ on the facet, which for discontinuous variables does not
have a unique definition. Summing over the domain gives

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
as in \citet{}, to take the average value of $\Pi$ across the facets,
$\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each
side of the facet. The brackets $\jump{q} = q^+-q^-$ denote the jump
in $q$ across the facet and the integral over the horizontal facets
vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous
across these facets so the jump is zero.

\subsection{Transport terms}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}^{HDiv}$, $\mathbb{V}^{L^2}$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable. The spatial order of the
finite element spaces also imposes requirements on the choice of
transport scheme. In particular, when the lowest order space
$\mathbb{V}^{L2}$ is chosen to be piecewise constant, we need to use
the recovered transport scheme described in
\citet{bendall2019recovered} to achieve second order convergence
overall. Here we describe the default transport schemes used in Gusto
for the next-to-lowest order sequence of function spaces (i.e. where
the lowest order space $\mathbb{V}^{L2}$ is chosen to be piecewise
linear and discontinuous between elements) and refer the reader to
\citet{bendall2019recovered} for details of the recovered schemes,
which are also implemented in Gusto.

\subsubsection{Transport of $\mathbb{V}^{L^2}$ fields}
The shallow water and compressible Euler equation sets both contain a
prognostic equation that is the continuity form of the transport
equation for a variable represented by a function that is fully
discontinuous between elements. This means that the divergence term is
not well defined. A stable discretiation of this equation can be
constructed by integrating that term by parts and taking the upwind
values on the element facets. Using the compressible Euler equation
notation, this gives, in each element $e$ with boundary $\partial e$:

\begin{equation}
  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\tilde{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
\end{equation}
Summing over the domain gives
\begin{equation}
  \int_\Omega \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_\Omega \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_\Gamma \jump{\phi\MM{u}}\tilde{\rho} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
\end{equation}
where $\Omega$ is the domain, $\Gamma$ denotes the element facets,
$\tilde{\rho}$ denotes the value of $\rho$ on the upwind side of the
facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
across the facet.

\subsubsection{Transport of $\mathbb{V}_\theta$ fields}
The compressible Boussinesq and compressible Euler equation sets both
contain a prognostic equation that is the advective form of the
transport equation for a variable that is represented by a function
that is continuous in the vertical direction but discontinuous in the
horizontal direction. This means that the gradient of the field
required in the transport term is not well defined. A stable
discretisation of this equation can be constructed by using a
combination of the above upwinding strategy in the discontinuous
horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
method in the continuous vertical dirction. The SUPG method replaces
the finite element test function $\gamma$ with $\gamma + \tau
\MM{u}\cdot\nabla\gamma$ where $\tau$ is a stabilisation parameter,
although here we just require the vertical part of the gradient term. This
has the effect of suppressing spurious oscillation associated with the
continuous finite element approximation of the transport term. The
stabilisation parameter $\tau$ depends on the spatial resolution and
** and is set to be ** by default in Gusto, following \citet{}.

We first integrate by parts in each column $C$ of the mesh, taking the
upwind value of the field, indicated by a $\tilde{}$, on the vertical
facets of the mesh, denoted by $\Gamma_v$. Using the notation of the
compressible Euler equations, this gives

\begin{align}
  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x &= -\int_C \nabla\cdot(\gamma\MM{u})\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}\tilde{\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta, \\
  &= \int_C \gamma\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}(\tilde{\theta}-\theta_{int})\diff S, \quad \forall \gamma \in \mathbb{V}_\theta,
\end{align}
where we have integrated by parts a second time to move the
differential operator from $\gamma$ back to $\theta$ since we need to
replace the test function $\gamma$ with $\gamma+\tau w\gamma_z$, where
$w=\MM{u}\cdot\hat{\MM{z}}$ is the vertical component of the
velocity. This gives a second boundary term that now contains the
value of $\theta$ in the interior of the column, denoted by
$\theta_{int}$. Summing over the domain and applying the SUPG method
in the vertical gives
\begin{equation}
  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
\end{equation}
Note that, as explained in \citet{Colin}, the test function should
also be replaced in the time derivative term for consistency **

\subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
